Stage-Splitting a Modal Language

### 2014-07-24 paper/semantics.tex
informal discussion

To understand how this behavior can be achieved, consider evaluation of the stage-\bbtwo\ term below, which contains nested $\next$ and $\prev$ blocks:
\begin{lstlisting} 
prev{
  let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
}
\end{lstlisting}
Our goal is to evaluate away all of the stage-\bbone\ portions of this code, but there is more than one reasonable way to do this.
One option, taken by \cite{davies96}, is to treat \verb|next{1+2}| as a value for the purposes of stage-\bbone~evaluation, 
consequently duplicating it during substitution for {\tt x}.  The two $\prev$ constructs eliminate the $\next$ wrappers,
leaving the contained stage-\bbtwo\ code in place.
This produces:
\begin{lstlisting}
(1+2) * (1+2) * 7
\end{lstlisting}

>>>>

\TODO clean up the next few paragraphs; unclear how much of it should go in the
paper

To understand the difficulty of staged evaluation, consider the following
example:
\begin{lstlisting}
#2 (next {1+2}, 3+4)
\end{lstlisting}
This is a stage-\bbone\ expression of type $\rmint$; the pair is a stage-\bbone\
expression of type $(\fut\rmint)\times\rmint$. A conventional call-by-value
evaluation strategy demands that we evaluate both components of the pair to
values before we project from the pair. However, at stage \bbone\ we are not
allowed to evaluate under the \verb|next|; on the other hand, \verb|next {1+2}|
is not a value, in the sense that it requires additional computation steps.

Intuitively, the solution is to say that \verb|next {1+2}| is a value \emph{at
stage \bbone}, even though once we proceed to stage-\bbtwo\ evaluation, it will
no longer be a value. Therefore, we evaluate the pair to
\begin{lstlisting}
#2 (next {1+2}, 7)
\end{lstlisting}
and get \verb|7| as the answer.

To complicate matters, consider a situation where we must substitute a stage
\bbone\ value into some other expression. In the stage-\bbtwo\ term
\begin{lstlisting} 
prev{
  let x = (next {1+2}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
}
\end{lstlisting}
our stage-\bbone\ evaluation will clearly yield a term which is not fully
reduced---since \verb|1+2| is needed to compute the end result, but we cannot
reduce it to \verb|3| until stage \bbtwo.

On the other hand, if we simply treat \verb|(next {1+2}, 7)| as a value, then we
will duplicate the \verb|1+2| computation in the term, and in fact we will
evaluate it twice, in each computation of \verb|#1 (next {1+2}, 7)|.

Our solution is to hoist these residuals, etc.

\vspace{2em}
\TODO old stuff below here. merge with above text:

Our goal is to evaluate away all of the stage-\bbone\ portions of this code, but there is more than one reasonable way to do this.
One option, taken by \cite{davies96}, is to treat \verb|next{1+2}| as a value for the purposes of stage-\bbone~evaluation, 
consequently duplicating it during substitution for {\tt x}.  The two $\prev$ constructs eliminate the $\next$ wrappers,
leaving the contained stage-\bbtwo\ code in place.
This produces:
\begin{lstlisting}
(1+2) * (1+2) * 7
\end{lstlisting}
